A Computational Account of Default-Mode Function\\by Control Theory and Reinforcement Learning

### 2016-07-16 paper.tex
more fixups

  \paragraph{Free-utility functional.}
    Let $(U, P)$ be a conjugate pair at temperature $\alpha > 0$  on a measurable space $(\Omega, \mathcal F)$. Given another probability measure $P'$ on the same space, define it's \textit{free utility} as
    \begin{equation}
      J(P'|U, P) = \mathbb E_{P'}(U) + \alpha \mathbb H(P'),
    \label{eq:free_u}
    \end{equation}
    where $\mathbb H(P') := -\sum_{\omega \in \Omega}P'(\omega)\log(P'(\omega))$ is the \textit{entropy} of $P'$ (measured in the Naperian base $e \approx 2.73$). It's not difficult to establish the upper bound
  \begin{equation}
    J(P'|U,P) \le J(P|U,P) = \sum_{\omega \in \Omega}U(\omega) =: U(\Omega).
  \end{equation}
  In particular, if $P$ is the Gibbs measure corresponding to $U$, then the upper bound above reduces to the \textit{log-partition function}
  \begin{equation}
    J(P'|U,P) \le U(\Omega) = -\alpha \log(Z_{U}(\alpha)).
    \end{equation}

  \paragraph{The free-energy / utility principle (of Friston ?).}
  We are now in shape to introduce the notion of free-energy for model transitions, and a variational principle for optimizing it. Consider thus an initial system described by a conjugate pair $(U_{\text{ini}}, P_{\text{ini}})$ at temperature $\alpha > 0$. We want to transform this to a new model by adding constraints represented by the utility function $\Delta U$.  The resulting system has final utility $U_{\text{fin}} = U_{\text{ini}} + \Delta U$. The difference in free-utility is then
  \begin{equation}
    \Delta J_{(U_{\text{ini}}, P_{\text{ini}}) \rightarrow P_{\text{fin}}} := J_{\text{fin}} - J_{\text{ini}} = \underbrace{\mathbb E_{P_{\text{fin}}}(\Delta U)}_{\textbf{accuracy}} + \alpha \underbrace{D_{\text{KL}}(P_{\text{ini}}\|P_{\text{fin}})}_{\textbf{complexity}},
    \label{eq:free}
  \end{equation}
  where $D_{\text{KL}}(P_{\text{ini}}\|P_{\text{fin}})$ is the Kullback-Leibler divergence.
  In the above formula, we've extensively used the fact that $(U_{\text{ini}}, P_{\text{ini}})$ is a congugate pair and so $U_{\text{ini}}(\omega) \equiv \alpha \log(P_{\text{ini}}(\omega))$ by virtue of \eqref{eq:boltzmann}.
  The two terms in \eqref{eq:free} (accuracy or expected gain in utility, and the complexity of the transition) can be viewed as dertiminants of bounded rational decision-making. They formalize
a trade-off between an expected utility $\Delta U$ (first term) and
the information cost of transforming Pi
into $P_{\text{fin}}$ (second
term). In this interpretation $P_{\text{ini}}$ represents an initial choice
probability or policy, which includes the special case of the
uniform distribution where the decision-maker has initially no
preferences between the different choices. The probability measure $P_{\text{fin}}$ is the final choice probability that we are looking for since it
considers the utility constraint U∗
that we want to optimize. We can then formulate a variational principle for bounded rationality in the probabilities $P_{\text{fin}}(\omega)$
\begin{equation}
P^*_{\text{fin}} := \underset{P_{\text{fin}}}{\text{argmax }} \Delta J_{(U_{\text{ini}}, P_{\text{ini}}) \rightarrow P_{\text{fin}}}
  \end{equation}

@@

Two limit cases are worth considering.
\begin{itemize}
\item \textbf{Low-temperature regime $\alpha \approx 0$:} Here $\Delta J_{(U_{\text{ini}}, P_{\text{ini}}) \rightarrow P_{\text{fin}}} \approx \mathbb E_{P'_{\text{fin}}}(\Delta U)$, and so it's optimal to take $P^*_{\text{fin}}(\omega) \equiv \text{dirac}(\omega - \omega^*)$, where $\omega^* := \argmax_{\omega \in \Omega}U(\omega)$. This corresponds to unbounded rational decision-making, in which the cost of transition / problem-solving is completely disregarded.
  \item \textbf{High-temperature regime $\alpha \rightarrow +\infty$:} In this limiting case, it's optimal to take $P^*_{\text{fin}}(\omega) \equiv P_{\text{ini}}(\omega)$, i.e the change is so costly that it's optimal to maintain the current choice probabilities.
  \end{itemize}
\section{Experimental Results}
\paragraph{Serial versus parallel structure discovery and classification.}

>>>>

  \paragraph{Free-utility functional.}
    Let $(U, P)$ be a conjugate pair at temperature $\alpha > 0$  on a measurable space $(\Omega, \mathcal F)$. Given another probability measure $P'$ on the same space, define it's \textit{free utility} as
    \begin{equation}
      J(P'|U, P) = \mathbb E_{P'}(U) + \alpha \mathbb H(P'),
    \label{eq:free_u}
    \end{equation}
    where $\mathbb H(P') := -\sum_{\omega \in \Omega}P'(\omega)\log(P'(\omega))$ is the \textit{entropy} of $P'$ (measured in the Naperian base $e \approx 2.73$). It's not difficult to establish the upper bound
  \begin{equation}
    J(P'|U) \le J(P|U) = \sum_{\omega \in \Omega}U(\omega) =: U(\Omega).
  \end{equation}
  In particular, if $P$ is the Gibbs measure corresponding to $U$, then the upper bound above reduces to the \textit{log-partition function}
  \begin{equation}
    J(P'|U) \le U(\Omega) = -\alpha \log(Z_{U}(\alpha)).
    \end{equation}

  \paragraph{The free-energy / utility principle (of Friston ?).}
  We are now in shape to introduce the notion of free-energy for model transitions, and a variational principle for optimizing it. Consider thus an initial system described by a conjugate pair $(U_{\text{ini}}, P_{\text{ini}})$ at temperature $\alpha > 0$. We want to transform this to a new model by adding constraints represented by the utility function $\Delta U$.  The resulting system has final utility $U_{\text{fin}} = U_{\text{ini}} + \Delta U$. The difference in free-utility is then
  \begin{equation}
    \Delta J_{(U_{\text{ini}}, P_{\text{ini}}) \rightarrow P_{\text{fin}}} := J_{\text{fin}} - J_{\text{ini}} = \underbrace{\mathbb E_{P_{\text{fin}}}(\Delta U)}_{\textbf{accuracy}} -  \underbrace{\alpha D_{\text{KL}}(P_{\text{ini}}\|P_{\text{fin}})}_{\textbf{complexity}},
    \label{eq:free}
  \end{equation}
  where
  \begin{equation}
    D_{\text{KL}}(P_{\text{fin}}\|P_{\text{ini}}) := \sum_{\omega \in \Omega}\P_{\text{fin}}(\omega)\log(\P_{\text{fin}}(\omega)/\P_{\text{ini}}(\omega))
  \end{equation}
  is the Kullback-Leibler divergence, and represents the information cost (measured in energy units) of changing the initial system.
In the above formula, we've extensively used the fact that $(U_{\text{ini}}, P_{\text{ini}})$ is a congugate pair and so $U_{\text{ini}}(\omega) \equiv \alpha \log(P_{\text{ini}}(\omega))$ by virtue of \eqref{eq:boltzmann}.
  The two terms in \eqref{eq:free} (accuracy or expected gain in utility, and the complexity of the transition) can be viewed as dertiminants of bounded rational decision-making. They formalize
a trade-off between an expected utility $\Delta U$ (first term) and
the information cost of transforming Pi
into $P_{\text{fin}}$ (second
term). In this interpretation $P_{\text{ini}}$ represents an initial choice
probability or policy, which includes the special case of the
uniform distribution where the decision-maker has initially no
preferences between the different choices. The probability measure $P_{\text{fin}}$ is the final choice probability that we are looking for since it
considers the utility constraint U∗
that we want to optimize. We can then formulate a variational principle for bounded rationality in the probabilities $P_{\text{fin}}(\omega)$
\begin{equation}
P^*_{\text{fin}} := \underset{P_{\text{fin}}}{\text{argmax }} \Delta J_{(U_{\text{ini}}, P_{\text{ini}}) \rightarrow P_{\text{fin}}}
  \end{equation}

@@

Two limit cases are worth considering.
\begin{itemize}
\item \textbf{Low-temperature regime $\alpha \approx 0$:} Here $\Delta J_{(U_{\text{ini}}, P_{\text{ini}}) \rightarrow P_{\text{fin}}} \approx \mathbb E_{P'_{\text{fin}}}(\Delta U)$, and so it's optimal to take
  $$P^*_{\text{fin}}(\omega) \equiv \text{dirac}(\omega - \omega^*) = \begin{cases}1, &\mbox{if }\omega = \omega^*,\\0, &\mbox{ otherwise,}\end{cases}$$
  where $\omega^* := \argmax_{\omega \in \Omega}U(\omega)$. This corresponds to unbounded rational decision-making, in which the cost of transition / problem-solving is completely disregarded.
  \item \textbf{High-temperature regime $\alpha \rightarrow +\infty$:} In this limiting case, it's optimal to take $P^*_{\text{fin}}(\omega) \equiv P_{\text{ini}}(\omega)$, i.e the change is so costly that it's optimal to maintain the current choice probabilities.
  \end{itemize}
\section{Experimental Results}
\paragraph{Serial versus parallel structure discovery and classification.}
